Unimodular covers of 3-dimensional parallelepipeds and Cayley sums

### 2023-06-22 unimodular-covers-CombTheory-final.tex
new claim for final argument

Let $f_q$ be the primitive lattice (affine) functional vanishing on $q$ and $f_p$ the one vanishing on $p$. We assume that $f_q(p_1) < f_q(p_2)$ and $f_p(q_1) < f_p(q_2)$.
Let $w=\area(p+q) \geq 2$, where $\area$ denotes the area normalized to a fundamental domain. In what follows, the width of a functional $f$ on a set $S$, denoted $\width_f(S)$ is defined as the difference $\sup_{x\in S} f(x)-\inf_{x\in S} f(x)$.
Then primitiveness of $p$, $q$, $f_p$ and $f_q$ implies that:
\begin{align*}
w=&\width_{f_q}(p + \vecline q )=\width_{f_q}(p) = f_q(p_2) - f_q(p_1) =\\
=&\width_{f_p}(q +\vecline p ) = \width_{f_p}(q)= f_p(q_2) - f_p(q_1).
\end{align*}

%
Since we can perform without loss of generality respective lattice translations to $P$ and to $Q$, we assume that the segments $p$ and $q$ intersect in their relative interiors and that $f_q(p_1)=-1$. That is,
\[
-1=f_q(p_1) < 0 < f_q(p_1), \quad
f_p(q_1) < 0 < f_p(q_2).
\]
In particular, $p$ must be in the interior of $q +\vecline p$ and $q$ in the interior of $p +\vecline q$
Our assumption also implies that $p_1$ is the unique lattice point with $f_q(x)=-1$ in $q +\vecline p$. Similarly, the unique lattice point in the strip with $f_q(x)=1$ is  $q_1+q_2 -p_1$.

@@

The last two claims can be summarized as saying that in the pictures $b$, $t$ and $r$ have positive slope. Observe that this implies that $q$ is not in the boundary of $Q$ and $p \neq r$, so both $P$ and $Q$ are full dimensional.

@@

We are now ready to show a contradiction. Since the normal fan of $Q$ refines that of $P$, $Q$ must contain an edge parallel to $r$ on the right. Let $r'$ be a primitive segment along this edge, so that $r'$ is a translated copy of $r$.
 Let $r_1'$ and $r_2'$ be its endpoints.

We consider two cases, depending on whether $q_1+q_2-p_1$ lies ``to the right'' or ``to the left'' of $p_1$. That is, whether $f_p(q_1+q_2-p_1)$ is positive or negative. %For illustration use the left part of Figure~\ref{fig:claim4} in which $f_p(q_1+q_2-p_1)=0$, and in which both arguments work:

\begin{itemize}
\item \underline{$f_p(q_1+q_2-p_1)> 0$}. In this case the segment $r$ in the boundary of $P$ separates $p$ from  $q_1+q_2-p_1$.
Let $r''_1$ and $r''_2$ be the intersection points of the line containing $r$ with $b$ and $t$ respectively. We have that
$r'=[r'_1,r'_2]$ is parallel to $[r''_1,r''_2]$ but shorter, so that
\[
w= \width_{f_q}(p) \le \width_{f_q}(r) = \width_{f_q}(r') \le  f_q(r''_2) -f_q(r''_1), 
\]
where the first inequality comes from \eqref{eq:widthp}.
But on the other hand
\[
f_q(r''_2) < f_q(q_1+q_2-p_1) = 1, \quad f_q(r''_1) > f_q(p_1) =-1,
\]
which leads to the contradiction $w<2$.

\item \underline{$f_p(q_1+q_2-p_1)\ge 0$}.
Consider the segment $s$ contained in $r_1' + \vecline p$ with endpoints $s_1=r_1'$ and $s_2$ on the line spanned by $t$. The width of $s$ with respect to $g$ is $g(s_2)-g(s_1) <1$, because $s$ is shorter than $Q \cap p$, which is strictly contained between the consecutive parallel lattice lines through $p_1$ and $q_2$ and through $q_1$ and $q_1+q_2-p_1$ where $g$ is constant.

We now observe that $g(r'_2) < g(s_2)$, because $r'_2$ lies to the right of $s_2$ ($r$ has positive slope) and below the line spanned by $t$, and $g$ decreases moving to the right along $t$ (\Cref{claim:g}). Thus $g(r_2)-g(r_1) = g(r'_2)-g(r'_1) < g(s_2)-g(s_1) <1$. Since $g(r_2)$ and $g(r_1)$ are integers this leads to $g(r_2)\le g(r_1)$, in contradiction with \Cref{claim:g}.
%Since $t$ separates $q_1$ and $q_1+q_2-p_1$ and $g$ decreases from $t_1$ to $t_2$ (by \Cref{claim:g}), the inequality $g(x)< g(d')$ holds on $Q\cap V_2$, and in particular for $r_2'$. Since $r_2'$ is a lattice point, $g(r_2')\leq g(d)= g(r_1')$, which contradicts \Cref{claim:g}).
\end{itemize}
\end{proof}

>>>>

Let $f_q$ be the primitive lattice (affine) functional vanishing on $q$ and $f_p$ the one vanishing on $p$, taking their signs so that $f_q(p_1) < f_q(p_2)$ and $f_p(q_1) < f_p(q_2)$.
Let $w=\area(p+q) \geq 2$, where $\area$ denotes the area normalized to a fundamental domain. In what follows, the width of a functional $f$ on a set $S$, denoted $\width_f(S)$ is defined as the difference $\sup_{x\in S} f(x)-\inf_{x\in S} f(x)$.
Then primitiveness of $p$, $q$, $f_p$ and $f_q$ implies that:
\begin{align*}
w=&\width_{f_q}(p + \vecline q )=\width_{f_q}(p) = f_q(p_2) - f_q(p_1) =\\
=&\width_{f_p}(q +\vecline p ) = \width_{f_p}(q)= f_p(q_2) - f_p(q_1).
\end{align*}

%
Since we can perform without loss of generality respective lattice translations to $P$ and to $Q$, we assume that 
$p_1$ lies on the line $\{f_q=-1\}$ and in the strip $q +\vecline p$. It  must then lie in the interior of the strip, or otherwise we would have that $p_2$ equals one of  $q_1$ or $q_2$ and $w=1$. That is, we have that $p$ and $q$ intersect in their relative interiors and that
\[
-1=f_q(p_1) < 0 < f_q(p_1), \quad
f_p(q_1) < 0 < f_p(q_2).
\]
%In particular, $p$ must be in the interior of $q +\vecline p$ and $q$ in the interior of $p +\vecline q$
Our assumption also implies that $p_1$ is the unique lattice point with $f_q(x)=-1$ in $q +\vecline p$. Similarly, the unique lattice point in the strip with $f_q(x)=1$ is  $q_1+q_2 -p_1$.

@@

The last two claims can be summarized as saying that in the pictures $b$, $t$ and $r$ have positive slope and that the slope of $b$ is greater or equal than that of $t$. Observe that this implies that $q$ is not in the boundary of $Q$ and $p \neq r$, so both $P$ and $Q$ are full dimensional.

@@

Let $r'=[r'_1,r'_2]$ be the segment collinear with $r$ and with endpoints $r'_1$ and $r'_2$ lying respectively on the lines containing $b$ and $t$. ($r'_1$ must in fact lie in $b$, but $r'_2$ may be outside $t$).

\begin{claim}
%\paco{new claim. Note I changed the roles of $r'$ and $r''$, since (the new) $r''$ is only used inside the claim}
\label{claim:r}
$r'$ has at least the same length as $r$. 
\end{claim}

\begin{proof}
Since the normal fan of $Q$ refines that of $P$, the boundary of $Q$ must contain an edge parallel to $r$ on the right. Let $r''$ be a primitive segment along this edge. Then $r''$ is a translated copy of $r$, and parallel to $r'$.

Now, $r''$ is contained also between the lines  along $b$ and $t$, and is to the right of $r'$ (because $r$, hence $r'$, has $q_2$ on its right, and $r''$ has $q_2$ on its left). Our assumption that the lines along $b$ and $t$ are either parallel or meet in $v_2$ implies the inequality.
\end{proof}

We are now ready to show a contradiction. We consider two cases, depending on whether $q_1+q_2-p_1$ lies ``to the right'' or ``to the left'' of $p_1$. That is, whether $f_p(q_1+q_2-p_1)$ is positive or negative. %For illustration use the left part of Figure~\ref{fig:claim4} in which $f_p(q_1+q_2-p_1)=0$, and in which both arguments work:

\begin{itemize}
\item \underline{$f_p(q_1+q_2-p_1)> 0$}. In this case the line containing $r$ and $r'$ separates $p$ from  $q_1+q_2-p_1$.
This implies
\[
f_q(r'_2) < f_q(q_1+q_2-p_1) = 1, \quad f_q(r'_1) > f_q(p_1) =-1,
\]
while Claims~\ref{claim:r} and~\ref{claim:width} imply that $\width_{f_q}(r')\ge \width_{f_q}(r)  \ge w\ge 2$.

\item \underline{$f_p(q_1+q_2-p_1)\le 0$}.
We observe that $g(r_2) - g(r_1)$ is positive (and integer) because $r$ has positive slope and separates $p_1$ from $q_2$. But Claim~\ref{claim:r} gives
\[
g(r_2) - g(r_1) \le g(r'_2) - g(r'_1) 
\]
which is in this case strictly smaller than $g(q_1+q_2-p_1) - g(p_1) =1$.
\qedhere
\end{itemize}
\end{proof}
